Pushdown Control-Flow Analysis via Refunctionalization (Functional Pearl)

### 2018-03-10 paper.tex
update

In ANF, an expression is an atomic expression or a @let@ expression.
The restriction is all the function applications must be administrated in a @let@ expression,
and then bounded to a variable name under the current environment.
Both the operator and operand of function applications should be an atomic expression.
The atomic expressions $ae$ are either a variable, or a literal @lambda@ term, which
both can be evaluated in a single step.

@@

TODO

@@

@@

To run (analyze) a program, as CESK machines, we first use function @inject@ to construct
the initial state given the program. 
Note that the initial continuation store has a built-in mapping that maps continuation address
@Halt@ to empty set of continuations.
We also provide an empty program contour as initial time.
\begin{lstlisting}
def inject(e: Expr): State = 
  State(e, Map(), 
        Store[BAddr, Storable](Map()), 
        Store[KAddr, Cont](Map(Halt -> Set())), 
        Halt, List())
\end{lstlisting}

@@

simply trun evaluation contexts into higher-order continuations.
We establish pushdown control-flow, but this is not computable.

\begin{lstlisting}
case class Config(e: Expr, env: Env, store: BStore, time: Time)
case class ValStore(ds: Set[Storable], store: BStore, time: Time)
type Cache = Map[Config, ValStore]

type Continue = Function2[ValStore, Cache, ValStore]

def tick(s: Config): Time = (s.e::s.time).take(k)

def aval(e: Expr, env: Env, bstore: BStore, time: Time, cache: Cache, continue: Continue): ValStore = { 
  val config = Config(e, env, bstore, time)
  if (cache.contains(config)) {
    return continue(cache(config), cache)
  }   
  val newTime = tick(config)
  e match {
    case Let(x, App(f, ae), e) if isAtomic(ae) =>
      val closures = aeval(f, env, bstore).toList

      def nd(closures: List[Storable], accds: Set[Storable],
             accbstore: BStore, acctime: Time, cache: Cache): ValStore  = { 
        closures match {
          case Nil =>
            val baddr = allocBind(x, acctime)
            val newEnv = env + (x -> baddr)
            val newStore = accbstore.update(baddr, accds)
            aval(e, newEnv, newStore, acctime, cache, (vs: ValStore, accCache: Cache) => {
              val newCache = accCache + (Config(e, newEnv, newStore, acctime) -> vs) 
              continue(vs, newCache)
            })  
          case Clos(Lam(v, body), env_c)::rest =>
            val baddr = allocBind(v, newTime)
            val newEnv = env_c + (v -> baddr)
            val newStore = accbstore.update(baddr, aeval(ae, env, bstore)) //TODO: which store?
            aval(body, newEnv, newStore, newTime, cache, (vs: ValStore, accCache: Cache) => {
              val newCache = accCache + (Config(body, newEnv, newStore, newTime) -> vs) 
              nd(rest, vs.ds++accds, accbstore, acctime, newCache)
            })  
        }   
      }

      nd(closures, Set(), bstore, newTime, cache)
    case ae if isAtomic(ae) =>
      val ds = aeval(ae, env, bstore)
      continue(ValStore(ds, bstore, newTime), cache)
  }   
}

def done: Continue = (vs: ValStore, cache: Cache) => { vs }                                                                                                                                                                                                 
def analyze(e: Expr): ValStore = aval(e, Map(), Store[BAddr, Storable](Map()),
                                      List(), Map[Config, ValStore](), done)
\end{lstlisting}

@@

\begin{lstlisting}
case class Config(e: Expr, env: Env)
case class ValStore(ds: Set[Storable], store: BStore, time: Time)
type Cache = Store[Config, ValStore]
type RetType = (Set[ValStore], Cache)
type Continue = Function5[Set[ValStore], BStore, Time, Cache, Cache, RetType]

def tick(e: Expr, oldtime: Time): Time = (e::oldtime).take(k)

def aval(e: Expr, env: Env, bstore: BStore, time: Time,
         inCache: Cache, _outCache: Cache, continue: Continue): RetType = {
  val config = Config(e, env)
  if (_outCache.contains(config)) {
    return continue(_outCache(config), bstore, List(), inCache, _outCache)
  }
  val outCache = _outCache.update(config, inCache.getOrElse(config, Set()))
  val newTime = tick(e, time)
  e match {
    case Let(x, App(f, ae), e) if isAtomic(ae) =>
      val closures = aeval(f, env, bstore).toList
      def nd(closures: List[Storable], accds: Set[Storable], accbstore: BStore, acctime: Time,
              accInCache: Cache, accOutCache: Cache): RetType  = closures match {
        case Nil =>
          // TODO: If we use `newTime` instead of `acctime`, is it m-CFA?
          val baddr = allocBind(x, acctime)
          val newEnv = env + (x -> baddr)
          val newStore = accbstore.update(baddr, accds)
          aval(e, newEnv, newStore, acctime, accInCache, accOutCache,
            (vs: Set[ValStore], accStore: BStore, acctime: Time, accInCache: Cache, accOutCache: Cache) => {
              val newOutCache = accOutCache.update(Config(e, newEnv), vs)
              continue(vs, accStore, acctime, accInCache, newOutCache)
            })
        case Clos(Lam(v, body), env_c)::rest =>
          val baddr = allocBind(v, newTime)
          val newEnv = env_c + (v -> baddr)
          val newStore = accbstore.update(baddr, aeval(ae, env, bstore)) //TODO: which store?
          aval(body, newEnv, newStore, newTime, accInCache, accOutCache,
            (vs: Set[ValStore], accStore: BStore, acctime: Time, accInCache: Cache, accOutCache) => {
              val newOutCache = accOutCache.update(Config(body, newEnv), vs)
              val vsds = vs.map(_.ds).foldLeft(Set[Storable]())(_ ++ _)
              nd(rest, vsds++accds, accStore, acctime, accInCache, newOutCache)
            })
      }
      nd(closures, Set(), bstore, newTime, inCache, outCache)
    case ae if isAtomic(ae) =>
      val ds = aeval(ae, env, bstore)
      continue(Set(ValStore(ds, bstore, newTime)), bstore, newTime, inCache, outCache)
    }
}

def done: Function2[Expr, Env, Continue] = (e: Expr, env: Env) => {
  (vs: Set[ValStore], store: BStore, time: Time, inCache: Cache, outCache: Cache) => {
    val newOutCache = outCache.update(Config(e, env), vs)
    (vs, outCache)
  }
}

def analyze1(e: Expr, cache: Cache): RetType = {
  aval(e, Map(), Store[BAddr, Storable](Map()), List(), cache, cache, done(e, Map()))
}

def analyze(e: Expr): Set[ValStore] = {
  def iter(cache: Cache): RetType = {
    val (vs, newCache) = analyze1(e, cache)
    if (cache == newCache) (vs, newCache)
    else iter(newCache)
  }
  val (vs, cache) = iter(Store[Config, ValStore](Map()))
  val summarized = cache.map.values.map(_.map(_.store).foldLeft(Store[BAddr, Storable](Map()))(_.join(_))).foldLeft(Store[BAddr, Storable](Map()))(_.join(_))
  vs
}
\end{lstlisting}

@@

\subsubsection{Computable Solutions}

In the rest of this section, we take a closer look at the P4F which is the
state-of-art techniques to accomplish perfect call-stack precision on small-step
AAMs, and Abstracting Defnitional Interpreters (ADI) that adpots another 
approach to pushdown control flow analysis.

@@

\citeauthor{gilray2016pushdown} proposed a continuation address allocator for 
small-step abstracting abstract machines that is both simple to implement and
computational inexpensive. 
Based on the AAM we presented in Section \ref{aam}, the only change in code is when 
allocating a continuation address, not only
keep track of the entry-point expression of callee, but also hold the target environment.
The target environment is the binding environment extended from closure's environment
(namely @val new_env = env_c + (v -> baddr)@). All the rest things remain the same.
The following code shows the implementation:

>>>>

In ANF, an expression is an atomic expression or a @let@ expression.
The restriction is all the function applications must be administrated in a @let@ expression,
and then bounded to a variable name under the current environment.
Both the operator and operand of function applications should be an atomic expression.
The atomic expressions $ae$ are either a variable, or a literal @lambda@ term, which
both can be evaluated in a single step.
We also assume that the program is globally alpha-renamed, i.e., all the
variable names are unique.

@@

A histroy of last $k$ call sites is served as time information in
$k$-call-sensitive analysis.
The history is represented as a list of expressions and embedded
in the allocated addresses.

Every time when we transfer to next states, we use function @tick@ to
fresh the time and use this new time when allocating addresses. The function
@tick@ returns the $k$ front-most expressions given the current state and its
time history.

@@

If we instantiate $k$ to be $0$, the history degenerates to an empty list,
then we obtain a monovariant analysis, which
not differentiates values at different call sites. In this case, the address
space collapses to the space of variable names.
Note that regarding to the ambiguitiy in
$k$-CFA\cite{Gilray:2016:ACP:2951913.2951936},
here the code actually implements a call+return sensitivity.

@@

To run (analyze) a program, as the CESK machines, we first use function @inject@ to construct
the initial state given the program. 
Note that the initial continuation store has a built-in mapping that maps continuation address
@Halt@ to empty set of continuations.
We also provide an empty program contour as initial time.
\begin{lstlisting}
def inject(e: Expr): State = 
  State(e, Map(), 
        Store[BAddr, Storable](Map()), 
        Store[KAddr, Cont](Map(Halt -> Set())), 
        Halt, List())
\end{lstlisting}

@@

Transform to Extended Continuation-Passing Style (ECPS), continuation and meta-continuation.
simply trun evaluation contexts into higher-order continuations.
We establish pushdown control-flow, but this is not computable.

\begin{lstlisting}
case class VS(vals: Set[Storable], time: Time, store: BStore)
type MCont = VS => VS
type Cont  = MCont => (VS => VS)

def aval(e: Expr, env: Env, store: BStore, time: Time, cont: Cont, rcont: MCont): VS = {
  val new_time = (e::time).take(k)

  e match {
    case Let(x, App(f, ae), e) =>
      val closures = atomicEval(f, env, store).toList.asInstanceOf[List[Clos]]

      def ndCont(closures: List[Clos])(r: MCont)(vs: VS): VS = {
        closures match {
          case Nil => r(vs)
          case Clos(Lam(v, body), c_env)::rest =>
            val baddr = allocBind(v, vs.time)
            val new_env = c_env + (v -> baddr)
            val new_store = vs.store.update(baddr, atomicEval(ae, env, vs.store))
            aval(body, new_env, new_store, new_time, ndCont(rest), (vs1: VS) => {
                    r(VS(vs.vals ++ vs1.vals, vs1.time, vs1.store))
                  })
        }
      }
      def mtCont(vs: VS): VS = {
        val baddr = allocBind(x, vs.time)
        val new_env = env + (x -> baddr)
        val new_store = vs.store.update(baddr, vs.vals)
        aval(e, new_env, new_store, vs.time, cont, rcont)
      }

      ndCont(closures)(mtCont)(VS(Set(), new_time, store))

    case ae if isAtomic(ae) =>
      val ds = atomicEval(ae, env, store)
      cont(rcont)(VS(ds, new_time, store))
  }
}

def analyze(e: Expr) = aval(e, Map(),
                            Store[BAddr, Storable](Map()),
                            List(),
                            (r: MCont) => (vs: VS) => vs,
                            (vs: VS) => vs)

\end{lstlisting}

@@

\begin{lstlisting}
\end{lstlisting}

@@

@@

\citeauthor{gilray2016pushdown} proposed a continuation address allocator for 
small-step abstracting abstract machines that is both simple to implement and
computational inexpensive. 
Based on the AAM we presented in Section \ref{aam}, the only change in code is
not only keeping track of the entry-point expression of callee, but also holding the target environment
when allocating a continuation address.
The target environment is the binding environment extended from closure's environment
(namely @val new_env = env_c + (v -> baddr)@). All the other part of the code remain the same.
The following code shows the implementation:
